---
title: "L07 Layers"
subtitle: "Data Visualization (STAT 302)" 
author: "Arturo Fonseca"
pagetitle: "L07 Arturo Fonseca"

format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-fold: show
    link-external-newwindow: true
    fig-dpi: 300

execute:
  warning: false

from: markdown+emoji 
reference-location: margin
citation-location: margin
---

## Overview

The goal of this lab is to explore more plots in `ggplot2` and continue to leverage the use of various layers to build complex and well annotated plots. 

## Datasets

We'll be using the `tech_stocks.rda` dataset which is already in the `/data` subdirectory in our **data_vis_labs** project. 

We have a new dataset, `NU_admission_data.csv`, which will need to be downloaded and added to our `/data` subdirectory. 

We will also be using the `mpg` dataset which comes packaged with `ggplot2` --- use `?ggplot2::mpg` to access its codebook.

```{r}
#| label: load-pkgs-data

# load package(s)

# load datasets

```

## Exercise 1

Using `mpg` and the `class_dat` dataset created below, recreate the following graphic as precisely as possible in two different ways.

*Hints:*

- Transparency is 0.6
- Horizontal spread is 0.1
- Larger points are 5
- Larger points are "red"

```{r}
#| label: ex-1
#| eval: false

# additional dataset for plot
class_dat <- mpg |>
  group_by(class) |>
  summarise(
    n = n(),
    mean_hwy = mean(hwy),
    label = str_c("n = ", n, sep = "")
  )
```

### Plot 1 -- using `mean_hwy`

::: {.callout-tip icon=false}

## Solution

:::

### Plot 2 -- not using `mean_hwy`

::: {.callout-tip icon=false}
## Solution

YOUR SOLUTION HERE

:::

## Exercise 2

Using the `perc_increase` dataset derived from the `tech_stocks` dataset, recreate the following graphic as precisely as possible.

*Hints:*

- Hex color code `#56B4E9`
- Justification of 1.1
- Size is 5

```{r}
#| label: ex-2
#| eval: false

# percentage increase data
perc_increase <- tech_stocks |>
  arrange(desc(date)) |>
  distinct(company, .keep_all = TRUE) |>
  mutate(
    perc = 100 * (price - index_price) / index_price,
    label = str_c(round(perc), "%", sep = ""),
    company = fct_reorder(factor(company), perc)
  )
```

::: {.callout-tip icon=false}
## Solution

YOUR SOLUTION HERE

:::

## Exercise 3

:::{.callout-warning}

## Warning

Some thoughtful data wrangling will be needed and it will be demonstrated in class --- **Do not expect a video**.

:::

Examine the data and the plot provided in `undergraduate-admissions-statistics.pdf` --- this pdf was collected from the [Northwestern Data Book webpage](https://www.adminplan.northwestern.edu/ir/data-book/). As you can see they have overlaid two plot types on one another by using dual y-axes. 

There is one major error they make with the bars in their graphic. **Explain what it is.**

::: {.callout-tip icon=false}
## Solution

YOUR SOLUTION HERE

:::

Using `NU_admission_data.csv`^[Data is taken from the pdf. The file includes a few extra years.], create **two separate plots** that display the same information instead of trying to put it all in one single plot --- **stack them** using `patchwork` or `cowplot`. 

Which approach do you find communicates the information better, the single dual y-axes plot or the two separate plot approach? **Why?**

::: {.callout-tip icon=false}
## Solution

YOUR SOLUTION HERE

:::

*Hints:*

- Form 4 datasets (helps you get organized, but not entirely necessary): 
    - 1 that has bar chart data, 
    - 1 that has bar chart label data, 
    - 1 that has line chart data, and 
    - 1 that has line chart labels 
- Consider using `ggsave()` to save the image with a fixed size so it is easier to pick font sizes.

::: {.callout-tip icon=false}
## Solution

YOUR SOLUTION HERE

:::
