---
title: "L08 Coords"
subtitle: "Data Visualization (STAT 302)"
author: "Arturo Fonseca"
pagetitle: "L08 Arturo Fonseca"
date: today

format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-fold: show
    link-external-newwindow: true
    fig-dpi: 300

execute:
  warning: false

from: markdown+emoji 
reference-location: margin
citation-location: margin
---


## Overview

The goal of this lab is to use an alternative coordinate system in `ggplot2` to build a plot. 

## Dataset

We'll be using the `mod_nba_adv_stats_2023_24.txt` dataset --- add to your project's `/data` subdirectory. The `mod_nba_adv_stats_2023_24_codebook.txt` provides a quick description of the variables in the dataset --- suggest adding it to the `/data` subdirectory as well. 

```{r}
#| label: load-pkgs-data

# load package(s)
library(tidyverse)

# load data
nba_dat <- read_delim("data/mod_nba_adv_stats_2023_24.txt")
```

## Exercise 

<!-- See the html instructions since they are complete with images and are the official instructions. -->

Using the `mod_nba_adv_stats_2023_24.txt` dataset we will try to recreate/approximate the plot type featured in the [FiveThirtyEight](http://fivethirtyeight.com/) article [Kawhi Leonard Is The Most Well-Rounded Elite Shooter Since Larry Bird](http://fivethirtyeight.com/features/kawhi-leonard-is-the-most-well-rounded-elite-shooter-since-larry-bird/) for any player of your choice for the 2023-2024 season. 

:::{.callout-note icon=false}

## Goal

Replicate the graphic below as closely as possible --- see instructions for image.

:::

### Start with data wrangling

When data wrangling we will need a helper function for creating the quartiles for players. Additionally, it will be useful to exclude players that played less than 10 games or played less than 5 minutes a game. That is, we only want to include "qualified" players.

```{r}
#| label: helper-function

# Returns quartile rank
quartile_rank <- function(x = 0:99) {
  # Set quartile
  quart_breaks <- c(
    -Inf,
    quantile(x,
      probs = c(.25, .5, .75),
      na.rm = TRUE
    ),
    Inf
  )
  cut(x = x, breaks = quart_breaks, labels = FALSE)
}
```

:::{.callout-important}
**Add comments to** the code below where indicated. The added comments should concisely describe what the following line(s) of code do in the data wrangling process  
:::

::: {.callout-tip icon=false}

## Solution

```{r}
#| label: data-wrangling

# Graphical data
nba_graph_dat <- nba_dat |> 
  # Exclude players that played less than 10 games or played less than 5 minutes a game
  filter(g >= 10, mp / g >= 5) |> 
  # Bin each percentage into their respective quartiles, 0%-25% -> 1, and so on...
  mutate(
    ts_quant = quartile_rank(ts_percent),
    trb_quant = quartile_rank(trb_percent),
    dbpm_quant = quartile_rank(dbpm),
    ast_quant = quartile_rank(ast_percent),
    usg_quant = quartile_rank(usg_percent)
  ) |> 
  # Only get the actual player and columns we made
  select(player, contains("_quant")) |> 
  # Making it tidy!
  pivot_longer(
    cols = -player, 
    names_to = "variable", 
    values_to = "value"
    ) |> 
  # Sort alphabetically by name
  arrange(player)
```

:::

### Build the plot

*Hints:*

- Image width is 8 in, height is 10 in --- use code chunk yaml
- center the image output --- use code chunk yaml
- shrink size down for display to 50% width --- use code chunk yaml
- All added text, except title, is done with `annotate()`
- set `axis.text` to blank in theme
- size 3 and 5 could be useful

::: {.callout-tip icon=false}
## Solution

```{r}
#| fig-width: 8
#| fig-height: 10
#| out-width: 50%
#| fig-align: "center"

filter(nba_graph_dat, player == "Anthony Edwards") |>
  ggplot(aes(variable, value)) +
  geom_col(width = 1,
           color = "black",
           fill = "#E38893") +
  geom_hline(yintercept = 1:4, linetype = "dotted") +
  scale_x_discrete(
    name = NULL,
    limits = c(
      "ts_quant",
      "usg_quant",
      "dbpm_quant",
      "trb_quant",
      "ast_quant"
    ),
    labels = NULL
  ) +
  scale_y_continuous(
    name = NULL,
    limits = c(NA, 5),
    breaks = NULL,
    labels = NULL
  ) +
  annotate(
    "text",
    x = 1:5,
    y = 5,
    label = c(
      "TRUE\nSHOOTING",
      "USAGE\nRATE",
      "DEFENSIVE\nBPM",
      "REBOUND\nRATE",
      "ASSIST\nRATE"
    ),
    size = 5
  ) +
  annotate(
    "text",
    x = 3,
    y = 1:4,
    label = c("1st-25th", "25th-50th", "50th-75th", "75th-100th"),
    vjust = -1,
    size = 3
  ) +
  coord_polar() +
  labs(title = "Anthony Edwards\n(2024)") +
  theme(axis.text = element_blank())
```

:::

## Challenge

:::{.callout-important icon=false}

Challenge is optional for all students, but we recommend trying them out!

:::

<!-- See the html instructions since they are complete with images and are the official instructions. -->

Continue the work of trying to recreate the graphic from the article. Your final plot should look as follows:

::: {.callout-tip icon=false}

```{r}
#| fig-width: 8
#| fig-height: 10
#| out-width: 50%
#| fig-align: "center"

filter(nba_graph_dat, player == "Anthony Edwards") |>
  ggplot(aes(variable, value)) +
  geom_col(width = 1,
           color = "black",
           fill = "#E38893") +
  geom_hline(yintercept = 1:4, linetype = "dotted") +
  scale_x_discrete(
    name = NULL,
    limits = c(
      "ts_quant",
      "usg_quant",
      "dbpm_quant",
      "trb_quant",
      "ast_quant"
    ),
    breaks = NULL,
    labels = NULL
  ) +
  scale_y_continuous(
    name = NULL,
    limits = c(NA, 5),
    breaks = NULL,
    labels = NULL
  ) +
  annotate(
    "text",
    x = 1:5,
    y = 5,
    label = c(
      "TRUE\nSHOOTING",
      "USAGE\nRATE",
      "DEFENSIVE\nBPM",
      "REBOUND\nRATE",
      "ASSIST\nRATE"
    ),
    size = 5
  ) +
  annotate(
    "text",
    x = 3,
    y = 1:4,
    label = c("1st-25th", "25th-50th", "50th-75th", "75th-100th"),
    vjust = -1,
    size = 3
  ) +
  coord_polar() +
  labs(title = "Anthony Edwards\n(2024)") +
  theme(axis.text = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 20))
```

:::